Optimal design of experiment

### 2024-09-19 Article_Design_of_Experiment-Chamomile.tex
19.09

@@

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
This study investigates the process of chamomile oil extraction from flowers. A parameter-distributed model consisting of a set of partial differential equations is used to describe the governing mass transfer phenomena in a solid-fluid environment under supercritical conditions using carbon dioxide as a solvent. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. The physical properties of the solvent are estimated from the Peng-Robinson equation of state. The empirical correlations used in the model are based on the set of laboratory experiments performed under multiple constant operating conditions: $30 - 40^\circ C$, $100 - 200$ bar, and $3.33-6.67 \cdot 10^{-5}$ kg/s. The model-based design of the experiments technique is applied in this work to improve the precision of the correlation parameters. The mass flow rate and the inlet temperature are used as decision variables to minimizing the volume of the confidence
ellipsoid. As a result, the optimal control problem is numerically solved to obtain the controls, which maximises the Fisher information contained in the yield curve.

@@

\section{Conclusions} \label{CH: Conclusion}

@@

>>>>

\usepackage{nomencl}
\makenomenclature

\usepackage{etoolbox}
\renewcommand\nomgroup[1]{%
	\item[\bfseries
	\ifstrequal{#1}{A}{Latin symbols}{%
		\ifstrequal{#1}{B}{Greek symbols}{%
			\ifstrequal{#1}{C}{Abberivations}{
	}}}%
	]}

@@

% ===================================================
% ABSTRACT
% ===================================================
\begin{abstract}
This study investigates the process of chamomile oil extraction from flowers. A parameter-distributed model consisting of a set of partial differential equations is used to describe the governing mass transfer phenomena in a solid-fluid environment under supercritical conditions using carbon dioxide as a solvent. The concept of quasi-one-dimensional flow is applied to reduce the number of spatial dimensions. The flow is assumed to be uniform across any cross-section, although the area available for the fluid phase can vary along the extractor. The physical properties of the solvent are estimated from the Peng-Robinson equation of state. The empirical correlations used in the model are based on the set of laboratory experiments performed under multiple constant operating conditions: $30 - 40^\circ C$, $100 - 200$ bar, and $3.33-6.67 \cdot 10^{-5}$ kg/s. A model-based design of experiments with D-optimality criterion is applied to improve the precision of the correlation parameters by designing a new dynamic experiment. The mass flow rate and inlet temperature are used as decision variables to maximize the Fisher Information embedded in the yield curve with respect to the empirical correlations.

@@

\section{Conclusions} \label{CH: Conclusion}
This paper introduces the D-optimal experimental design (D-OED) as a model-based method for finding an experimental procedure in chemical engineering problems. D-OED forms the objective function, which can be interpreted as minimising the volume of uncertainty ellipsoid. Compared to the classical formulation of D-OED, a penalty term was introduced to the cost function to prevent infeasibly rapid changes in operating conditions or bang-bang type control. The method is demonstrated through a case study on supercritical extraction. This work aims to design an experiment to improve the precision of the correlation for $D_i$. While the original experiments were conducted under constant operating conditions, this work looks for an experimental procedure with dynamically changing operating conditions. In such a case, the D-OED solution can also be used for model validation.

The analysis was repeated for multiple pressure values, and the obtained optimal profiles of inlet temperature and mass flow rate were similar across all cases. The resulting yield curves exhibit wavy patterns due to the relationship between the Hessian and the curvature of a multi-variable function. The analysis of the yield curves and scatter plots reveals that applying the same experimental strategy at different operating conditions results in varying amounts of information being collected. Consequently, it is concluded that the operating conditions should always be carefully chosen to conduct the most informative experiments. This is because the physical properties of CO$_2$ vary less at pressures far from the critical point, meaning that the same set of controls has less impact on the system at high pressures than at low pressures.

@@

\newpage

\nomenclature[A]{\(P\)}{Pressure}
\nomenclature[A]{\(T\)}{Temperature}
\nomenclature[A]{\(T^{in}\)}{Inlet temperature}
\nomenclature[A]{\(T^{out}\)}{Outlet temperature}
\nomenclature[A]{\(F\)}{Mass flow rate}
\nomenclature[A]{\(u\)}{Superfical velocity}
\nomenclature[A]{\(v\)}{Linear velocity}
\nomenclature[A]{\(z\)}{Spatial direction}
\nomenclature[A]{\(e\)}{Internal energy}
\nomenclature[A]{\(e^y\)}{Output error}
\nomenclature[A]{\(h\)}{Enthalpy}
\nomenclature[A]{\(t\)}{Time}
\nomenclature[A]{\(t_f\)}{Total extraction time}
\nomenclature[A]{\(t_0\)}{Inital extraction time}
\nomenclature[A]{\(A\)}{Total cross section of the bed}
\nomenclature[A]{\(A_f\)}{Cross section of the bed accupied by the fluid}
\nomenclature[A]{\(c_f\)}{Concentration of solute in fluid phase}
\nomenclature[A]{\(c_{f0}\)}{Inital concentration of solute in fluid phase}
\nomenclature[A]{\(c_f^*\)}{Concentration of solute at the solid-fluid interface}
\nomenclature[A]{\(c_s\)}{Concentration of solute in solid phase}
\nomenclature[A]{\(c_{s0}\)}{Inital concentration of solute in solid phase}
\nomenclature[A]{\(c_s^*\)}{Concentration of solute at the solid-fluid interface}
\nomenclature[A]{\(c_p\)}{Concentration of solute in the core of a pore}
\nomenclature[A]{\(c_{pf}\)}{Concentration of solute in the pore opening}
\nomenclature[A]{\(D_e^M\)}{Axial diffusion coefficeint}
\nomenclature[A]{\(r_e\)}{Mass transfer kinetic term}
\nomenclature[A]{\(L\)}{Length of fixed bed}
\nomenclature[A]{\(j\)}{Objective function}
\nomenclature[A]{\(j_{ml}\)}{Maximum Likelihood}
\nomenclature[A]{\(S\)}{Sensitivity equations}
\nomenclature[A]{\(Q\)}{Weighting matrix}
\nomenclature[A]{\(R\)}{Control cost matrix}
%\nomenclature[A]{\(\textbf{G}\)}{Augumented system}
%\nomenclature[A]{\(\dot{S}\)}{Time derivative of sensitivity equations}
%\nomenclature[A]{\(\bar{S}\)}{Sensitivity matrix}
%\nomenclature[A]{\(\bar{J}_x\)}{Jacobian matrix with respect to the state space}
%\nomenclature[A]{\(\bar{J}_\Theta\)}{Jacobian matrix with respect to the paramters}
%\nomenclature[A]{\(\mathcal{L}\)}{Log-likelihood function}
\nomenclature[A]{\(l\)}{Characterisitic diemsinon of particles}
\nomenclature[A]{\(D_i\)}{Interanl diffusion coefficient}
\nomenclature[A]{\(D_i^R\)}{Reference value of interanl diffusion coefficient}
\nomenclature[A]{\(r\)}{Particle radius}
\nomenclature[A]{\(k_p\)}{Volumetric partition coefficient}
\nomenclature[A]{\(k_m\)}{Mass partition coefficient}
\nomenclature[A]{\(y\)}{Extraction yield}
\nomenclature[A]{\(Y\)}{Yield measurment}
\nomenclature[A]{\(x\)}{State vector}
\nomenclature[A]{\(G\)}{Vector of discretized differential equations}
\nomenclature[A]{\(p\)}{Probaility disribution model}
\nomenclature[A]{\(R_e\)}{Reynolds number}
\nomenclature[A]{\(\mathcal{F}\)}{Fisher information}

\nomenclature[B]{\(\rho_f\)}{Fluid density}
\nomenclature[B]{\(\Phi\)}{Bed porosity}
\nomenclature[B]{\(\rho_s\)}{Bulk density of solid}
\nomenclature[B]{\(\mu\)}{Sphericity coefficient}
\nomenclature[B]{\(\pi\)}{Probability desnsity}
\nomenclature[B]{\(\gamma\)}{Decaying function}
\nomenclature[B]{\(\Upsilon\)}{Decay coefficient}
\nomenclature[B]{\(\Theta\)}{Paramter space}
\nomenclature[B]{\(\theta\)}{vector of unknown parameters}
\nomenclature[B]{\(\epsilon\)}{Unobervable error}
\nomenclature[B]{\(\sigma\)}{Standard deviation}
\nomenclature[B]{\(\Sigma\)}{Covariance matrix}
\nomenclature[B]{\(\xi\)}{Vector of experimental condtions}
\nomenclature[B]{\(\Xi\)}{Matrix of experimental condtions}

\nomenclature[C]{BIC}{Broke-and-Intact Cell model}
%\nomenclature[C]{MLE}{Maximum Likelihood Estimation}
\nomenclature[C]{SFE}{Supercritical Fluid Extraction}
\nomenclature[C]{DoE}{Design of Experiment}
\nomenclature[C]{m-DoE}{Model-based Design of Experiment}
\nomenclature[C]{HBD}{Hot Ball Diffusion}
\nomenclature[C]{SC}{Shrinking Core}

\printnomenclature

### 2024-09-19 Sections/Optimal_design_of_experiment.tex
19.09

	\subsubsection{Fisher information}

	The Fisher information is a way of measuring the amount of information that an observable random variable carries about an unknown parameter of a distribution that models the random variable. The Fisher information is related to the second derivative (or the curvature) of the log-likelihood function with respect to the parameter. This relationship provides a measure of how "sensitive" the likelihood is to changes in the parameter value. The Fisher information matrix $\mathcal{F}$ can be calculated as follow.

@@

	If every experiment at time $t_i$ is independent and characterized by its own $\sigma_{t_i}$, then the Fisher information matrix can be presented in the more compact way:

@@

	In other words, the precision to which we can estimate $\Theta$ is limited by the Fisher information of the likelihood function. Based on the Cramer-Rao inequality, the Fisher information matrix can used to calculate the covariance matrices associated with maximum-likelihood estimates.

@@

	The optimal design of experiments (DOE) is a statistical concept that refers to the process of planning an experiment, which allow parameters to be estimated without bias and with minimum variance. Optimal design ensures that the experiment can provide the most informative data possible. This often involves balancing the study of main effects and interactions between factors. Moreover, by efficiently planning experiments, optimal design aims to reduce the overall resources required, such as time, materials, and manpower.

@@

	The cost function should describe the amount of information from an experiment. For that purpose it can be assumed a function $\kappa$ can be related to the Fisher information obtained an arbitrary operating conditions.

@@

	and choosing $Q = \textbf{I}_{n_\Theta}$ corresponds to the A-optimality cost function. An A-optimal experiment minimizes the sum of the squares of the lengths of the axes of asymptotic confidence ellipsoids. When $Q$ diagonal, with $[Q]_{ii} = 1/\Theta_i$ this corresponds to C-optimality, which is associated with the relative precision of estimates. Selecting $Q$ as a row vector results in C-optimality. Setting $Q = \textbf{I}_{n_\Theta}$ and $k = \infty$ leads to E-optimality, where the design maximizes the smallest eigenvalues of the Fisher information matrix, thereby minimizing the length of the largest axis of the asymptotic confidence ellipsoids. The most commonly used criterion is the D-optimality that involves $k = 0$, $Q = \textbf{I}_{n_\Theta}$, requiring the minimization of $\det \mathcal{F}^{-1}(\Theta, \Xi)$, or, equivalently, maximization of

>>>>

	\subsubsection{Fisher Information}

	The Fisher Information is a way of measuring the amount of information that an observable random variable carries about an unknown parameter of a distribution that models the random variable. The Fisher Information is related to the second derivative (or the curvature) of the log-likelihood function with respect to the parameter. This relationship provides a measure of how "sensitive" the likelihood is to changes in the parameter value. The Fisher Information matrix $\mathcal{F}$ can be calculated as follow.

@@

	If every experiment at time $t_i$ is independent and characterized by its own $\sigma_{t_i}$, then the Fisher Information matrix can be presented in the more compact way:

@@

	In other words, the precision to which we can estimate $\Theta$ is limited by the Fisher Information of the likelihood function. Based on the Cramer-Rao inequality, the Fisher Information matrix can used to calculate the covariance matrices associated with maximum-likelihood estimates.

@@

	The optimal design of experiments is a statistical concept that refers to the process of planning an experiment, which allow parameters to be estimated without bias and with minimum variance. Optimal design ensures that the experiment can provide the most informative data possible. This often involves balancing the study of main effects and interactions between factors. Moreover, by efficiently planning experiments, optimal design aims to reduce the overall resources required, such as time, materials, and manpower.

@@

	The cost function should describe the amount of information from an experiment. For that purpose it can be assumed a function $\kappa$ can be related to the Fisher Information obtained an arbitrary operating conditions.

@@

	and choosing $Q = \textbf{I}_{n_\Theta}$ corresponds to the A-optimality cost function. An A-optimal experiment minimizes the sum of the squares of the lengths of the axes of asymptotic confidence ellipsoids. When $Q$ diagonal, with $[Q]_{ii} = 1/\Theta_i$ this corresponds to C-optimality, which is associated with the relative precision of estimates. Selecting $Q$ as a row vector results in C-optimality. Setting $Q = \textbf{I}_{n_\Theta}$ and $k = \infty$ leads to E-optimality, where the design maximizes the smallest eigenvalues of the Fisher Information matrix, thereby minimizing the length of the largest axis of the asymptotic confidence ellipsoids. The most commonly used criterion is the D-optimality that involves $k = 0$, $Q = \textbf{I}_{n_\Theta}$, requiring the minimization of $\det \mathcal{F}^{-1}(\Theta, \Xi)$, or, equivalently, maximization of

### 2024-09-19 Sections/Results_DOE.tex
19.09

@@

>>>>

	\begin{comment}

@@

	\end{comment}

	To identify the global solution for Equations \ref{EQ:Formulation_2}, the optimization problem is solved multiple times, each run starting from a random initial solution sampled from a uniform distribution. Figure \ref{fig:scatter} compares the initial and final cost function values across multiple optimization runs for all cases. The solution with the lowest cost function value is considered the global solution for each case.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/scatter.png}	
		\caption{Initial vs final values of the cost function}
		\label{fig:scatter}
	\end{figure}

	From Figure \ref{fig:scatter}, it is evident that multiple local solutions exist. By analyzing locations of the clusters in Figure \ref{fig:scatter}, it can be concluded that experiments conducted near the critical point provide more information regarding the correlation than those conducted farther from it. The closer the pressure is to the critical point, the larger the deviations in the physical properties of $CO_2$ caused by changes in the controls, leading to greater variation in the Reynolds number, and consequently to more informative experiments.

	For each pressure case, the profiles with the lowest value of the objective function are further analysed. As shown in Figure \ref{fig:profiles_T}, all the inlet temperature profiles follow the same pattern. Initally, the system starts from the highest temperature (upper control bound) and gradually decreasing to lowest temperatures (lower control bound).

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/Profile_T.png}	
		\caption{Optimal inlet temperature profile}
		\label{fig:profiles_T}
	\end{figure}

	As presented in Figure \ref{fig:profiles_F}, in four out of five analysed cases, the corresponding mass flow rate profiles initially start at low values and gradually increase, reaching their maximum when the inlet temperature reaches its minimum. Later, the mass flow rates decrease to minimum values, and eventually, they rise again to reach their maximums at the final time. The only solution, which has different mass flow rate profile corresponds to pressure of 100 bar. The mass flow rate profile make the U-shape pattern, which starts the maximum value, then diminish to minimum and increases to the maximum value.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/Profile_F.png}	
		\caption{Optimal mass flow rate profiles}
		\label{fig:profiles_F}
	\end{figure}

	Figure \ref{fig:profiles_y} shows the predicted yield curves. The yield and scatter plots show that the most informative experiments do not necessarily correspond to those with the highest yield. This is because the variation in $CO_2$'s physical properties differs depending on the system's pressure. At higher pressures, the extraction rates are high enough that the amount of collected oils becomes almost the same, which makes it difficult to distinguish between experiments. The optimal yield profiles are characterized by a wavy pattern rather than the smooth exponential-like pattern observed under steady conditions. This behaviour is expected, as the determinant of the Hessian matrix, when evaluated at a critical point of a function, is equal to the Gaussian curvature of the function considered as a manifold. When the sum of the Hessian determinants is maximized, the number of critical points increases, leading to the wavy pattern observed in the yield curves.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=\columnwidth]{Figures/Results/yield.png}	
		\caption{Optimal yield profiles}
		\label{fig:profiles_y}
	\end{figure}
